---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
## unzip and read file
data_file <- unzip("./activity.zip")
activity <- read.csv(data_file)
```


## What is mean total number of steps taken per day?
```{r}
library(ggplot2)
total_steps <- tapply(activity$steps, activity$date, FUN=sum, na.rm=TRUE)
qplot(total_steps, xlab="total number of steps taken per day")
mean(total_steps, na.rm=TRUE)
median(total_steps, na.rm=TRUE)
```


## What is the average daily activity pattern?

```{r}
library(ggplot2)
averages <- aggregate(x=list(steps=activity$steps), by=list(interval=activity$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps")
```

On average across all the days in the dataset, the 5-minute interval that contains the maximum number of steps is

```{r}
averages[which.max(averages$steps),]
```

## Imputing missing values

```{r}
missing <- is.na(activity$steps)
# How many missing
table(missing)
```

Strategy for filling in all of the missing values in dataset.
Strategy is to fill all of the missing values with mean value for that 5-minute
interval.

```{r}

fill_value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled_data <- activity
# data set with no missing values
filled_data$steps <- mapply(fill_value, filled_data$steps, filled_data$interval)
```
Now using the filled dataset, create a histogram of the total number of steps taken each day and calculate the mean and median total number of steps.

```{r}
total_steps <- tapply(filled_data$steps, filled_data$date, FUN=sum)
qplot(total_steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total_steps)
median(total_steps)
```

Mean and median values came up higher after imputing missing data. 

## Are there differences in activity patterns between weekdays and weekends?
We will use filled-in dataset to find the day of the week for each measurement.

```{r}
day_of_measurement <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled_data$date <- as.Date(filled_data$date)
filled_data$day <- sapply(filled_data$date, FUN=day_of_measurement)
```

We will create a panel plot containing a time series plot  of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
averages <- aggregate(steps ~ interval + day, data=filled_data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```
